---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The `propr` package implements two related analyses: proportionality analysis and differential proportionality analysis. Proportionality is a compositional alternative to correlation analysis. It is introduced by Lovell et al., expounded by Erb & Notredame, and implemented by Quinn et al. (2017). Differential proportionality is a compositional alternative to differential abundance analysis. It is introduced by Erb et al. (2017), and discussed further by Quinn et al. (2019). Compositional data analysis and genomics is reviewed by Quinn et al. (2018).

If you use this software, please cite our work. We don't get paid to make software, but your citations allow us to negotiate time allocation for software maintence and development.

```{r}
citation("propr")
```

OK, now let's get started.

```{r, eval = FALSE}
counts <- matrix(rpois(20*50, 100), 20, 50)
group <- sample(c("A", "B"), size = 20, replace = TRUE)
devtools::install_github("tpq/propr")
library(propr)
```

## Proportionality

There are a few proportionality statistics available. Select one with the 'metric' argument.

```{r, eval = FALSE}
pr <- propr(counts, # rows as samples, like it should be
            metric = "rho", # or "phi", "phs", "cor", "vlr"
            ivar = "clr", # or can use "iqlr" instead
            alpha = NA, # use to handle zeros
            p = 100) # used by updateCutoffs
```

You can determine the "signficance" of proportionality using a built-in permutation procedure. It tells estimates the false discovery rate (FDR) for any cutoff. This method can take a while to run, but is parallelizable.

```{r, eval = FALSE}
updateCutoffs(pr,
              cutoff = seq(0, 1, .05), # cutoffs at which to estimate FDR
              ncores = 1) # parallelize here
```

Choose the largest cutoff with an acceptable FDR.

## Differential Proportionality

There are also a few differential proportionality statistics, but they all get calculated at once.

```{r, eval = FALSE}
pd <- propd(counts,
            group, # a vector of 2 or more groups
            alpha = NA, # whether to handle zeros
            weighted = TRUE, # whether to weigh log-ratios
            p = 100) # used by updateCutoffs
```

You can switch between the "disjointed" and "emergent" statistics.

```{r, eval = FALSE}
setDisjointed(pd)
```

```{r, eval = FALSE}
setEmergent(pd)
```

You can again permute an FDR with the `updateCutoffs` method. Alternatively, you can calculate an exact p-value for $\theta$ based on a F-test. This is handled by the `updateF` method.

```{r, eval = FALSE}
pd <- updateF(pd,
              moderated = FALSE, # moderate stats with limma-voom
              ivar = "clr") # used for moderation
```

## Getters

Both functions return S4 objects. This package includes several helper functions that work for both the `propr` and `propd` output. Most of the time, you would use `getResults`. You can select pairs greater or less than a certain size with the 'cutoff' argument.

```{r, eval = FALSE}
?getResults # get results in long-format
?getMatrix # get results as a square matrix
?getAdj # get an adjacency matrix
```

The vignettes describe some custom visualization methods.
