---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

## Quick start

Welcome to the `propr` GitHub page!

The bioinformatic evaluation of gene co-expression often begins with correlation-based analyses. However, this approach lacks statistical validity when applied to relative count data. This includes, for example, those biological data produced by microarray assays or high-throughput RNA-sequencing. This package provides a set of functions for measuring dependence between relative features using compositional data analysis. Specifically, this package implements two measures of *proportionality*, &phi; and &rho;, introduced in Lovell 2015 and expounded in Erb 2016. You can get started with `propr` by installing the most up-to-date version of this package directly from GitHub.

```{r, eval = FALSE}
library(devtools)
devtools::install_github("tpq/propr")
library(propr)
```

```{r, echo = FALSE}
library(propr)
```

The principal functions in `propr` include: (1) `phit`, for the calculation of &phi;, and (2) `perb`, for the calculation of &rho;. In the example below, we calculate proportionality for a simulated dataset, print the results as a proportionality matrix, then the index pairs of interest within the proportionality matrix. We refer you to the official package vignette for a comprehensive discussion of compositional data, proportionality, and everything this package has to offer.

```{r}
set.seed(12345)
N <- 10
data.absolute <- data.frame(a=(1:N), b=(1:N) * rnorm(N, 10, 0.1),
                            c=(N:1), d=(N:1) * rnorm(N, 10, 1.0))
data.relative <- data.absolute / colSums(data.absolute)
```

### Calculate &phi;

```{r, fig.show = "hold"}
phi <- phit(data.relative)
phi@matrix
phi05 <- phi["<", .05]
phi05@pairs
plot(phi05)
```

### Calculate &rho;

```{r, fig.show = "hold"}
rho <- perb(data.relative)
rho@matrix
rho99 <- rho[">", .99]
rho99@pairs
plot(rho99)
```

### References

1. Erb, I. & Notredame, C. 2016. How should we measure proportionality on relative gene expression data? Theory Biosci.

2. Lovell, D. et al. 2015. Proportionality: A Valid Alternative to Correlation for Relative Data. PLoS Comput Biol 11.
