---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Attention!!

If you have used `propr` previously, you will notice some changes. In version 5.0.0, I have completed a major revision of the code base to simplify the maintenance of `propr` going forward, including a restructure of back-end and front-end functionality. From the user's perspective, you will notice a few changes. First, all supporting visualization functions are gone. They were poorly implemented and not backwards compatible. Instead, you can use the unified `getResults` wrapper to pull data from `propr` and `propd` objects to pipe to `ggplot2` for visualization. Second, many experimental functions have been removed, with the remaining ones all sharing the prefix `run`. The core routines called by the `propr` and `propd` functions remain unchanged.

## Introduction

The `propr` package provides an interface for 4 distinct approaches to compositional data analysis (CoDA): proportionality, differential proportionality, partial correlation, and ratio analysis.

If you use this software, please cite our work. We don't get paid to make software, but your citations help us to negotiate support for software maintenance and development.

```{r}
citation("propr")
```

OK, now let's get started.

```{r, eval = FALSE}
counts <- matrix(rpois(20*50, 100), 20, 50)
group <- sample(c("A", "B"), size = 20, replace = TRUE)
devtools::install_github("tpq/propr")
library(propr)
```

## Proportionality

There are a few proportionality statistics available. Select one with the 'metric' argument.

```{r, eval = FALSE}
pr <- propr(counts, # rows as samples, like it should be
            metric = "rho", # or "phi", "phs", "cor", "vlr"
            ivar = "clr", # or can use "iqlr" instead
            alpha = NA, # use to handle zeros
            p = 100) # used by updateCutoffs
```

You can determine the "signficance" of proportionality using a built-in permutation procedure. It tells estimates the false discovery rate (FDR) for any cutoff. This method can take a while to run, but is parallelizable.

```{r, eval = FALSE}
updateCutoffs(pr,
              cutoff = seq(0, 1, .05), # cutoffs at which to estimate FDR
              ncores = 1) # parallelize here
```

Choose the largest cutoff with an acceptable FDR.

## Differential Proportionality

There are also a few differential proportionality statistics, but they all get calculated at once.

```{r, eval = FALSE}
pd <- propd(counts,
            group, # a vector of 2 or more groups
            alpha = NA, # whether to handle zeros
            weighted = TRUE, # whether to weigh log-ratios
            p = 100) # used by updateCutoffs
```

You can switch between the "disjointed" and "emergent" statistics.

```{r, eval = FALSE}
setDisjointed(pd)
```

```{r, eval = FALSE}
setEmergent(pd)
```

You can again permute an FDR with the `updateCutoffs` method. Alternatively, you can calculate an exact p-value for $\theta$ based on a F-test. This is handled by the `updateF` method.

```{r, eval = FALSE}
pd <- updateF(pd,
              moderated = FALSE, # moderate stats with limma-voom
              ivar = "clr") # used for moderation
```

## Getters

Both functions return S4 objects. This package includes several helper functions that work for both the `propr` and `propd` output.

```{r, eval = FALSE}
?getMatrix # get results as a square matrix
?getResults # get propr or propd results in long-format
?getRatios # get samples by ratios matrix
```

Use `getResults` to pipe to `ggplot2` for visualization.

## Partial Correlation

COMING SOON!!

## Ratio Methods

COMING SOON!!
