---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

## Quick start

Welcome to the `propr` GitHub page!

The bioinformatic evaluation of gene co-expression often begins with correlation-based analyses. However, this approach lacks statistical validity when applied to relative data. This includes, for example, biological count data generated by high-throughput RNA-sequencing, chromatin immunoprecipitation (ChIP), ChIP-sequencing, Methyl-Capture sequencing, and other techniques. This package implements several metrics for proportionality, including phi [Lovell et al (2015)] and rho [Erb and Notredame (2016)]. This package also implements several metrics for differential proportionality. Unlike correlation, these measures give the same result for both relative and absolute data.

```{r, eval = FALSE}
library(devtools)
devtools::install_github("tpq/propr")
library(propr)
```

```{r, echo = FALSE}
library(propr)
```

## Further reading

To learn more about proportionality, we refer the reader to the relevant literature.

```{r}
citation("propr")
```
