---
title: "Calculating the Proportionality Coefficient of Compositional Data with propr"
author: "Thomas Quinn"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## 

## Empiric Distributions

The proportionality metrics generated by the `phit` and `perb` functions do not permit an easy interpretation of "How proportional is proportional enough?". For this reason, we include an option to generate a null distribution of proportionality metrics through randomization of feature (e.g. gene) vectors. Effectively, this yields a kernel density function that estimates how likely `propr` would measure that degree of proportionality by chance alone. In its implementation, these empiric distributions consider only the absolute value of the proportionality metrics calculated by `perb`.

When using `phit` or `perb`, by setting the argument `iter` to a non-zero value, the `@pairs` slot of the resultant `propr` object will list an empiric p-value and FDR corrected p-value for each pair. The argument `iterSize` will specify how many features to include in each iteration. We emphasize here that these empiric distributions exist as a convenient guide for selecting features from ranked space and *not* as a means by which to form the basis of statistical hypothesis testing.

```{r, echo = TRUE}
set.seed(12345)

randomNum <- sample(1:1000, size = 2000 * 22, replace = TRUE)
counts <- matrix(randomNum, nrow = 2000, ncol = 22)
prop <- phit(counts, symmetrize = FALSE, iter = 50, iterSize = 500)

sum(prop$pval < .01) # 20,422 pairs show rare proportionality
sum(prop$fdr < .01) # ...but none survive FDR correction
```

## Visualizing Highly Proportional Pairs

Each feature (e.g. gene) belonging to a highly proportional data pair should show linearly correlated expression with one another across all subjects. The method `plot` provides a means by which to visually inspect whether this holds true. This function will plot *all* pairs in the `@pairs` slot. Therefore, we strongly recommend the user first subset the `propr` object before plotting. "Noisy" correlation between some feature pairs could suggest that the proportionality cutoff is too lenient.



## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
