---
title: "Weighted Disproportionality Update"
author: "Thomas Quinn"
date: "May 22, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Extensible theta typing

In addition to $\theta_d$ (for disjointed proportionality) and $\theta_e$ (for emergent proportionality), `propr` version 3.0.2 now addps support for $theta_f$ (equivalent to $1 - \theta_e$). I have chosen "f" (code name: fettered proportionality) for no reason other than to continue the alphabetical sequence.

```{r, fig.keep = "last"}
library(propr)
data(caneToad.counts)
keep <- apply(caneToad.counts, 2, function(x) sum(x > 40) >= 20)
ct <- caneToad.counts[, keep]
test <- ct[, sample(1:ncol(ct))[1:1000]]
group <- caneToad.groups
pd <- propd(test, group)
```

I have replaced the `setDisjointed` and `setEmergent` functions with the more useful `setActive` function. This new function (among other changes) will make it easier to add more theta variants in the future.

```{r}
pd.d <- setActive(pd, what = "theta_d")
head(pd.d@theta)
pd.e <- setActive(pd, what = "theta_e")
head(pd.e@theta)
pd.f <- setActive(pd, what = "theta_f")
head(pd.f@theta)
```

## Introducing weights

A new argument, `weighted`, will calculate $\theta_d$, $\theta_e$, and $\theta_f$ using `limma::voom` weights. I have written the weighted lrm and lrv calculations in C++, so they run extremely fast. See "tests/testthat/test-wt.R" for code validating that results from C++ implementation match those returned by the (slower) R implementation.

```{r}
pd.w <- propd(test, group, weighted = TRUE)
pd.dw <- setActive(pd.w, what = "theta_d") # just like unweighted
head(pd.dw@theta)
```

The `propd` object created by the `propd` function now contains two additional slots to track which theta type is under study. This is used by `updateCutoffs` to make it possible to permute FDR for *any* theta type *with or without* weights.

```{r, eval = FALSE}
pd.f <- updateCutoffs(pd.f, cutoff = seq(.5, .95, .05))
pd.dw <- updateCutoffs(pd.dw, cutoff = seq(.5, .95, .05))
```

Note that the `alpha` parameter will work for $\theta$,  $\theta_e$, and $\theta_f$. However, for now, trying to use the alpha transformation in addition to weights will trigger an error.

```{r, eval = FALSE}
propd(test, group, alpha = .01, weighted = TRUE) # throws error
```

## Visualization

All visualization functions work the same for $\theta_f$ and the weighted thetas. Out of curiosity, I generate all figures for $\theta_f$ and $\theta_w$.

```{r, fig.keep = "last"}
bowtie(pd.d, prompt = FALSE)
```

```{r, fig.keep = "last"}
bowtie(pd.f, prompt = FALSE)
```

```{r, fig.keep = "last"}
bowtie(pd.dw, prompt = FALSE)
```

```{r, fig.keep = "last"}
geiser(pd.d, cutoff = 1000, k = 5, prompt = FALSE)
```

```{r, fig.keep = "last"}
geiser(pd.f, cutoff = 1000, k = 5, prompt = FALSE)
```

```{r, fig.keep = "last"}
geiser(pd.dw, cutoff = 1000, k = 5, prompt = FALSE)
```

```{r, fig.keep = "last"}
gemini(pd.d, cutoff = 1000, k = 5, prompt = FALSE)
```

```{r, fig.keep = "last"}
gemini(pd.f, cutoff = 1000, k = 5, prompt = FALSE)
```

```{r, fig.keep = "last"}
gemini(pd.dw, cutoff = 1000, k = 5, prompt = FALSE)
```

```{r}
plot(pd.d, cutoff = 1000, prompt = FALSE)
plot(pd.f, cutoff = 1000, prompt = FALSE)
plot(pd.dw, cutoff = 1000, prompt = FALSE)
```
